

```markdown
---
title: "A tour of MicrobiomeExplorer"
author: "Your Name"
date: "`r format(Sys.time(), '%d %b %Y')`"
output: 
  rmarkdown::html_vignette:
    toc: true
    number_sections: false
vignette: >
  %\VignetteIndexEntry{A tour of MicrobiomeExplorer}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include=FALSE}
library(knitr)
opts_chunk$set(fig.align = "center", 
               out.width = "90%",
               fig.width = 6, fig.height = 5.5,
               dev.args=list(pointsize=10),
               collapse = TRUE,
               warning = FALSE)
set.seed(1) # for exact reproducibility
```

## Introduction

`MicrobiomeExplorer` is an R package designed to facilitate the analysis and visualization of microbiome datasets. It includes functions to calculate statistical metrics, normalize data, and create comprehensive heatmaps for exploratory data analysis. **This document provides a guide to using the MicrobiomeExplorer package (version 0.1.0)**.

To install **MicrobiomeExplorer**, run the following commands:

``` r
if (!require("devtools"))
  install.packages("devtools")
devtools::install_github("YourGitHubUsername/MicrobiomeExplorer", build_vignettes = TRUE)
library(MicrobiomeExplorer)
```

To list all functions available in the package:
``` r
ls("package:MicrobiomeExplorer")
```

To list all datasets available in the package:
``` r
data(package = "MicrobiomeExplorer")
```

<br>

## Using MicrobiomeExplorer

### Data Preparation

Before analyzing your microbiome data, ensure it is in the correct format: a matrix or dataframe with rows representing samples and columns representing taxa.

### Calculating Statistics

Use the `calculate_stats` function to compute extended statistics for your dataset:

``` r
# Example dataset
data(microbiome_example)

# Calculate statistics
stats <- MicrobiomeExplorer::calculate_stats(microbiome_example)
print(stats)
```

### Visualizing Data

Visualize your microbiome data with a heatmap using the `plot_microbiome_heatmap` function:

``` r
# Generate a heatmap
MicrobiomeExplorer::plot_microbiome_heatmap(microbiome_example, normalize = TRUE)
```

<br>

## Conclusion

The `MicrobiomeExplorer` package provides an easy-to-use set of tools for microbiome data analysis, allowing researchers to quickly calculate key statistics and visualize their data. For any further details, consult the function-specific help pages.

<br>

## Session Information

It's always a good practice to include the session information in vignettes.

```{r}
sessionInfo()
```
```

Remember to replace "Your Name" and "YourGitHubUsername" with your actual name and GitHub username. After saving this as an `.Rmd` file in the `vignettes` folder, you can build your vignettes using the `devtools::build_vignettes()` function.
